FATE summary results

### 2019-09-12 paper draft.Rmd
Changes to plots

```{r, echo=F}
GOA.ich <- readRDS("/Users/MikeLitzow/Documents/R/Fate/fate-ewi/GOA_ich_7.19.18.one.trend.rds") 
GOA.ich$summary
```

@@

plot.a <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("a) Ichthyoplankton loadings")

@@

plot.b <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") +
  theme(axis.title.x = element_blank()) + 
  ggtitle("b) Ichthyoplankton trend")

@@

ggplot(mod, aes(x=SST)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + xlab("Slope")

@@

```{r, echo=F}
GOA.zoop <- readRDS("/Users/MikeLitzow/Documents/R/Fate/fate-ewi/GOA_zoop_7.19.18.one.trend.rds") 
GOA.zoop$summary
```

@@

plot.d <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("d) Zooplankton loadings")

@@

plot.e <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") +
  theme(axis.title.x = element_blank()) + 
  ggtitle("e) Zooplankton trend")

@@

set.seed(99)
mod2 = trend_lm(rotated_modelfit = r, all_years = years,
  selected_years = years, climate_dat=cli_dat)

# get 95% CIs
x <- sort(mod2[,2])
print(names(mod2)[2])
print(paste("median: ", median(x)))
print(paste("LCI: ", x[(0.025*25000)]))
print(paste("UCI: ", x[(0.975*25000)]))

mod2$covar <- "SST"

ggplot(mod2, aes(x=covar, y=SST)) + 
  theme_bw() +
  geom_violin() + 
  geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + xlab("Slope")

@@

# Now the EBS results.

EBS.clim <- readRDS("/Users/MikeLitzow/Documents/R/Fate/fate-ewi/EBS_clim_6.19.18.rds") # read in EBS climate model

EBS.clim$summary
```

And the trends and loading plots.

# recover the years and names for plots!
ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))
keep <- grep("AKCLIM_EBS", ewidata$code)

ewidata <- ewidata[keep,]
sub_data = ewidata

rotated = rotate_trends(EBS.clim$best_model)

new.names <- c("Bottom.temp", "Spring.SST", "Winter.SST", "Summer.NW.wind", "Summer.SE.wind", "Winter.NW.wind", "Winter.SE.wind", "Ice.57.75N", "Ice.58.75N", "Ice.64.25N")

# loadings
pdf("ebs.clim.loadings.pdf", 6, 6)
print(plot_loadings(rotated, names = new.names, threshold=0.9))
dev.off()

# trends
pdf("ebs.clim.trend.pdf", 6,6)
print(plot_trends(rotated, years = as.numeric(colnames(Y))))
dev.off()
```

```{r, echo=F}
xx <- find_swans(rotated, threshold = 0.01, plot=F)
xx$year <- 1950:2016
xx <- filter(xx, below_threshold == T)
xx
```

Test for regime probability. Use n=3 maximum states to rule out a switch to a warmer state.

Regime results for Trend 1. The 2-regime model is best-supported.

## EBS biology results.

```{r, echo=F}
EBS.biol <- readRDS("/Users/MikeLitzow/Documents/R/Fate/fate-ewi/EBS_biol_6.19.18.rds") 
EBS.biol$summary
```

Now plot trends and loadings.

# reload the version of data used
# load updated ewidata!
ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))

# first, EBS biology
codes <- unique(ewidata$code)

# limit to EBS
EBS <- ewidata[ewidata$system == "EBS",]

# and start in 1975, when data beyond salmon are first available
EBS <- EBS[EBS$year>=1975,]

unique(EBS$code)

# drop clim
drop <- grep("AKCLIM", EBS$code)
EBS <- EBS[-drop,]

# drop depth
drop <- grep("DEPTH", EBS$code)
EBS <- EBS[-drop,]

# recover the years and names for plots!
sub_data = EBS

rotated = rotate_trends(EBS.biol$best_model)

new.names <- c("Alaska.place.lat", "Bristol.Bay.coho", "Bering.flounder.lat", "St.George.kitt.phen", "St.Paul.kitt.phen", "St.Paul.kitt.product", "com.eider.eggs/nest", "com.eider.hatch.date", "com.eider.success", "St.George.com.murre.phenol", "St.Paul.com.murre.phenol", "St.George.com.murre.product", "St.Paul.com.murre.product", "juv.cod.cpue", "juv.pollock.cpue", "flathead.sole.lat", "gl.gull.eggs/nest", "gl.gull.hatch.date", "gl.gull.success", "Grnlnd.turbot.lat", "Pac.cod.lat", "Pac.halibut.lat", "red.king.crab.lat", "rock.sole.lat", "shortf.eelpout.lat", "snow.crab.lat", "spec.eider.eggs/nest", "spec.eider.hatch.date", "spec.eider.success", "sturgeon.poacher.lat", "Tanner.crab.lat", "St.George.tb.murre.phenol", "St.Paul.tb.murre.phenol", "St.George.tb.murre.product", "St.Paul.tb.murre.product", "walleye.pollock.lat", "wattled.eelpout.lat", "yellowfin.sole.lat")

# loadings
pdf("ebs.biol.loadings.pdf", 6, 6)
print(plot_loadings(rotated, names = new.names, threshold=0.9))
dev.off()

```{r, echo=F}
# trends

pdf("ebs.biol.trend.pdf", 6,6)
print(plot_trends(rotated, years = as.numeric(colnames(Y))))
dev.off()
```
Test for regime probability. Use n=3 maximum states to rule out a switch to a warmer state.

Regime results for Trend 1. The 2-regime model is best-supported.

Next step is to examine relationships between SST and the EBS biology trend.

```{r, message=FALSE, results='hide', echo=F}

ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))
keep <- grep("AKCLIM_EBS", ewidata$code)
clim.var <- ewidata[keep,]  
keep <- grep("win.sst", clim.var$code)
clim.var <- clim.var[keep,]

win.sst <- tapply(clim.var$value, clim.var$year, identity)

years <- as.numeric(colnames(EBS.biol$best_model$data)) # these are the years for the biology model - 1975:2016

cli_dat <- data.frame(Year=as.numeric(names(win.sst)), SST=win.sst)

cli_dat$slope <- 1
cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
cli_dat$slope[cli_dat$Year %in% 2014:2016] <- 3
cli_dat$slope <- as.factor(cli_dat$slope)

set.seed(4)
mod = trend_lm(rotated_modelfit = r, all_years = 1975:2016,
  selected_years = 1975:2016, climate_dat=cli_dat)

names(mod)[2:4] <- c("1975-1988", "1989-2013", "2014-2016")
summary(mod)

melt.mod <- melt(mod, measure.vars=c("1975-1988", "1989-2013", "2014-2016"))
names(melt.mod)[3] <- "era"

pdf("Fig. x ebs.biol.on.sst.pdf", 6, 4)
ggplot(melt.mod, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="A) SST")
dev.off()
```

```{r, message=FALSE, results='hide', echo=F}

years <- as.numeric(colnames(EBS.biol$best_model$data)) # these are the years for the biology model - 1972:2016

# set up climate data
cli_dat <- data.frame(Year=pdo$YEAR,PDO=win.pdo)

cli_dat$slope <- 1
cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
cli_dat$slope[cli_dat$Year %in% 2014:2016] <- 3
cli_dat$slope <- as.factor(cli_dat$slope)

set.seed(99)
mod2 = trend_lm(rotated_modelfit = r, all_years = 1975:2016,
  selected_years = 1975:2016, climate_dat=cli_dat)

names(mod2)[2:4] <- c("1975-1988", "1989-2013", "2014-2016")

melt.mod <- melt(mod2, measure.vars=c("1975-1988", "1989-2013", "2014-2016"))
names(melt.mod)[3] <- "era"

pdf("Fig. x ebs.biol.on.pdo.pdf", 6, 4)
ggplot(melt.mod, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="B) PDO")
dev.off()

And the winter AO.

ao <- read.csv("winter.ao.csv")

win.ao <- ao$NDJFM

years <- as.numeric(colnames(EBS.biol$best_model$data)) # these are the years for the biology model - 1975:2016

# set up climate data
cli_dat <- data.frame(Year=ao$YEAR,AO=win.ao)

cli_dat <- filter(cli_dat, Year %in% years) %>% mutate(AO = arm::rescale(AO))

cli_dat$slope <- 1
cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
cli_dat$slope[cli_dat$Year %in% 2014:2016] <- 3
cli_dat$slope <- as.factor(cli_dat$slope)

Y = rnorm(length(cli_dat$slope))
cli_dat = as.data.frame(model.matrix(Y ~ Year + AO:slope, data=cli_dat))
# replace ':' in name for formula
names(cli_dat) = gsub(":","",names(cli_dat))
# remove named intercept
cli_dat = cli_dat[,-which(names(cli_dat)=="(Intercept)")]

set.seed(99)
mod3 = trend_lm(rotated_modelfit = r, all_years = 1975:2016,
  selected_years = 1975:2016, climate_dat=cli_dat)

names(mod3)[2:4] <- c("1975-1988", "1989-2013", "2014-2016")

summary(mod3)

for(i in 2:4){
x <- sort(mod3[,i])
print(names(mod3)[i])
print(paste("median: ", median(x)))
print(paste("LCI: ", x[(0.025*25000)]))
print(paste("UCI: ", x[(0.975*25000)]))
}

# and proportion > 0
for(i in 2:4){
x <- sort(mod3[,i])
print(names(mod3)[i])
print(paste("prop > 0: ", sum(x > 0)/length(x)))
}

melt.mod <- melt(mod3, measure.vars=c("1975-1988", "1989-2013", "2014-2016"))
names(melt.mod)[3] <- "era"

pdf("Fig. x ebs.biol.on.ao.pdf", 6, 4)
ggplot(melt.mod, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="B) AO")
dev.off()
```

```{r}
head(melt.mod)
restricted <- filter(melt.mod, era %in% c("1975-1988", "1989-2013"))
restricted <- melt.mod
png("ebs community response to ao.png", 6,4, units="in", res=300)
ggplot(data=restricted, aes(fill=era, x=era, y=value)) + geom_violin() + coord_flip() + ggtitle("Community response to Arctic Oscillation") +
  geom_hline(yintercept = 0) + xlab("Era") + ylab("Regression coefficient") + theme_gray() + theme(legend.position = c(0.8, 0.5), legend.title = element_blank())
dev.off()
```

>>>>

```{r, echo=F}
GOA.ich <- readRDS("/Users/MikeLitzow 1/Documents/R/Fate/fate-ewi/GOA_ich_7.19.18.one.trend.rds") 
GOA.ich$summary
```

@@

plot.a <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("a) Ichthyoplankton loadings") +
  theme(title = element_text(size=8))

@@

plot.b <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") +
  xlab("Year") +
  ggtitle("b) Ichthyoplankton trend") +
  theme(title = element_text(size=8))

@@

plot.c <- ggplot(mod, aes(x=SST)) +
  theme_bw() +
  geom_density(fill=cb[3]) + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept = 0, lty=2) + 
  xlab("Slope") + 
  ylab("Density") +
  ggtitle(("c) Ichthyoplankton slope on SST")) +
  theme(title=element_text(size=8))

@@

```{r, echo=F}
GOA.zoop <- readRDS("/Users/MikeLitzow 1/Documents/R/Fate/fate-ewi/GOA_zoop_7.19.18.one.trend.rds") 
GOA.zoop$summary
```

@@

plot.d <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("d) Zooplankton loadings") +
  theme(title = element_text(size=8))

@@

plot.e <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") + xlab("Year") +
  ggtitle("e) Zooplankton trend") +
  theme(title = element_text(size=8))

@@

plot.f <- ggplot(mod, aes(x=SST)) +
  theme_bw() +
  geom_density(fill=cb[3]) + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept = 0, lty=2) + 
  xlab("Slope") + 
  ylab("Density") +
  ggtitle(("f) Zooplankton slope on SST")) +
  theme(title = element_text(size=8))

@@

And combine into a single plot.

```{r}
png("combined short time series plots.png", 8,5, units="in", res=300)

ggpubr::ggarrange(plot.a, plot.b, plot.c, plot.d, plot.e, plot.f,
          ncol=3, widths=c(1, 1, 0.8), nrow=2)
